Sumarize and Explore Data by Shashank Verma
========================================================

```{r echo=FALSE, message=FALSE, warning=FALSE, set_directory}
# Set the working directory and list the files
setwd("C:/Users/shank/Desktop/DAND/P4")
```

```{r echo=FALSE, message=FALSE, warning=FALSE, packages}
# Load all of the packages that you end up using in your analysis in this code
# chunk.

# Notice that the parameter "echo" was set to FALSE for this code chunk. This
# prevents the code from displaying in the knitted HTML output. You should set
# echo=FALSE for all code chunks in your file, unless it makes sense for your
# report to show the code that generated a particular plot.

# The other parameters for "message" and "warning" should also be set to FALSE
# for other code chunks once you have verified that each plot comes out as you
# want it to. This will clean up the flow of your report.

library(ggplot2)
library(grid)
library(gridExtra)
library(GGally)
library(dplyr)
library(tidyr)
library(reshape)


```

```{r echo=FALSE, message=FALSE, warning=FALSE, Load_the_Data}
# Load the Data
wine <- read.csv("wineQualityReds.csv")
```

> The dataset is related to red variants of the Portuguese "Vinho Verde" wine. \
The inputs include objective tests (e.g. PH values) and the output is based on \
sensory data (median of at least 3 evaluations made by wine experts). Each \
expert graded the wine quality between 0 (very bad) and 10 (very excellent). \
Only physicochemical (inputs) and sensory (the output) variables are available.

> I will explore the wine quality and physicochemical properties of these \
wines. I want to know how the chemical properties influence the quality of \
red wines.

# Summary Statistics

```{r echo=FALSE, message=FALSE, warning=FALSE, Summary_Statistics}
names(wine)
```

```{r echo=FALSE, message=FALSE, warning=FALSE, Summary_Statistics_2}
str(wine)
```

```{r echo=FALSE, message=FALSE, warning=FALSE, Summary_Statistics_3}
summary(wine)
```

> There are 1599 observations of 13 numeric variables. \
X appears to be the unique identifier. quality is an ordered, categorical, \
discrete variable. From the literature, this was on a 0-10 scale, \
and was rated by at least 3 wine experts. The values ranged only from 3 to 8, \
with a mean of 5.6 and median of 6. All other variables seem to be continuous \
quantities (with the exception of the .sulfur.dioxide suffixes). \
From the variable descriptions, it appears that \
fixed.acidity ~ volatile.acidity and \
free.sulfur.dioxide ~ total.sulfur.dioxide may possibly be \
dependent, subsets of each other.


```{r echo=FALSE, message=FALSE, warning=FALSE, Transform_data}

# Transform quality from an integer to an ordered factor
wine$quality <- factor(wine$quality, ordered = T)

# create a rating varible
wine$rating <- ifelse(wine$quality < 5, 'bad', 
                      ifelse(wine$quality < 7, 'average', 'good'))
wine$rating <- ordered(wine$rating, levels = c('bad', 'average', 'good'))
```

# Univariate Plots Section

> Now, I'll draw quick histograms of all the features for showing the \
distributions of the variables.

```{r echo=FALSE, message=FALSE, warning=FALSE, Univariate_Plots_1}
fix_ac_plot = ggplot(aes(x = fixed.acidity), data = wine) +
  geom_histogram(stat = "count")

vol_ac_plot = ggplot(aes(x = volatile.acidity), data = wine) +
  geom_histogram(stat = "count")

cit_ac_plot = ggplot(aes(x = citric.acid), data = wine) +
  geom_histogram(stat = "count")

res_sug_plot = ggplot(aes(x = residual.sugar), data = wine) +
  geom_histogram(stat = "count")

chlor_plot = ggplot(aes(x = chlorides), data = wine) +
  geom_histogram(stat = "count")

free_sul_plot = ggplot(aes(x = free.sulfur.dioxide), data = wine) +
  geom_histogram(stat = "count")

tot_sul_plot = ggplot(aes(x = total.sulfur.dioxide), data = wine) +
  geom_histogram(stat = "count")

den_plot = ggplot(aes(x = density), data = wine) +
  geom_histogram(stat = "count")

ph_plot = ggplot(aes(x = pH), data = wine) +
  geom_histogram(stat = "count")

sul_plot = ggplot(aes(x = sulphates), data = wine) +
  geom_histogram(stat = "count")

al_plot = ggplot(aes(x = alcohol), data = wine) +
  geom_histogram(stat = "count")

qual_plot = ggplot(aes(x = quality), data = wine) +
  geom_histogram(stat = "count")

grid.arrange(fix_ac_plot, vol_ac_plot, cit_ac_plot, res_sug_plot, 
chlor_plot, free_sul_plot, tot_sul_plot, den_plot, ph_plot, 
sul_plot, al_plot, qual_plot, ncol = 3)

```

> pH is normally distributed, but fixed.acidity and citric.acid appear to \
be positively skewed. 

> Here are some quick box plots for all the numerical variables. We can \
visualize the outliers here.
 
```{r echo=FALSE, message=FALSE, warning=FALSE, Univariate_Plots_2}
meltData <- melt(wine[2:13])
ggplot(meltData, aes(factor(variable), value))+ 
  geom_boxplot() + facet_wrap(~variable, scale="free")
```

> We will draw log 10 plots of these variables to analyse further.

```{r echo=FALSE, message=FALSE, warning=FALSE, Univariate_Plots_3}
ggplot(wine, aes(x = fixed.acidity)) +
  geom_histogram() +
  scale_x_log10()

ggplot(wine, aes(x = volatile.acidity)) +
  geom_histogram() +
  scale_x_log10()

ggplot(wine, aes(x = citric.acid)) +
  geom_histogram() +
  scale_x_log10()
```

> On a log 10 scale, fixed.acidity and volatile.acidity appear to have a \
normal distribution, which makes sense since pH has a normal distribution. \
However, citric.acid does not appear to have a normal distribution.

```{r echo=FALSE, message=FALSE, warning=FALSE, Univariate_Plots_4}
length(subset(wine, citric.acid == 0)$citric.acid)
```

> After checking further, we find that the citric.acid contains 132 zero \
values. This can arise if the values were not entered.

```{r echo=FALSE, message=FALSE, warning=FALSE, Univariate_Plots_5}
p1 <- ggplot(aes(x = residual.sugar), data = wine) +
  geom_histogram()

p2 <- p1 + 
  scale_x_continuous(lim = c(0, quantile(wine$residual.sugar, 0.95))) +
  xlab('residual.sugar, 95th percentile truncated')

p3 <- p2 + scale_x_log10() + xlab('residual.sugar, log10')
grid.arrange(p1, p2, p3, ncol=1)

p1 <- ggplot(aes(x = chlorides), data = wine) +
  geom_histogram()

p2 <- p1 + 
  scale_x_continuous(lim = c(0, quantile(wine$chlorides, 0.95))) +
  xlab('chlorides, 95th percentile truncated')

p3 <- p2 + scale_x_log10() + xlab('chlorides, log10')
grid.arrange(p1, p2, p3, ncol=1)

p1 <- ggplot(aes(x = sulphates), data = wine) +
  geom_histogram()

p2 <- p1 + 
  scale_x_continuous(lim = c(0, quantile(wine$sulphates, 0.95))) +
  xlab('sulphates, 95th percentile truncated')

p3 <- p2 + scale_x_log10() + xlab('sulphates, log10')
grid.arrange(p1, p2, p3, ncol=1)

```

> Above are the plots for residual.sugar, chlorides and sulphates with \
the original plot, 95th percentile truncated and then the log 10 plots. \
We can get a better understanding of the distribution here.

# Univariate Analysis

> Below are the distributions of the features found in our dataset:

* Positively Skewed: fixed.acidity, citric.acid, free.sulphur.dioxide, \
total.sulphur.dioxide, sulphates, alcohol
* Normal: volatile.acidity, density, pH
* Long Tailed: residual.sugar, chlorides


### What is the structure of your dataset?
> The dataset consists of 1599 observations of 13 variables. quality is an \
ordered categorical variable and all the other variables are numerical.

### What is/are the main feature(s) of interest in your dataset?
> The main feature of interest here is the quality of wine. It ranges from \
3 to 8 in the dataset.

### What other features in the dataset do you think will help support your \
investigation into your feature(s) of interest?
> The multiple acidity variables can be related to each other and the quality \
of the wines. The alcohol and residual.sugar variables also look interesting.


### Did you create any new variables from existing variables in the dataset?
> I transformed the quality from an integer to an ordered factor and created \
a variable called rating to classify the wines as good, average or bad.


### Of the features you investigated, were there any unusual distributions? \
Did you perform any operations on the data to tidy, adjust, or change the form \
of the data? If so, why did you do this?
> I did not perform any operations to tidy the data, only changed the \
quality from an integer to an ordered factor and created a variable \
called rating to classify the wines as good, average or bad as this will help \
in better visualising the quality of wines.

# Bivariate Plots Section


```{r echo=FALSE, message=FALSE, warning=FALSE, Bivariate_Cors}
cor(
  wine %>%
    # first we remove unwanted columns
    dplyr::select(-X) %>%
    dplyr::select(-rating) %>%
    mutate(
      # now we translate quality to a number
      quality = as.numeric(quality)
    )
  )

```

> From the above correlations we can see that the quality is correlated with 

* alcohol (positive correlation)
* volatile.acidity (negative correlation)
* sulphates (positive correlation)
* citric.acid (positive correlation)

```{r echo=FALSE, message=FALSE, warning=FALSE, Bivariate_Plots_2}
get_bivariate_boxplot <- function(x, y, ylab) {
  return(ggplot(data = wine, aes(x = x, y = y)) + geom_boxplot() + ylab(ylab))
}

grid.arrange(get_bivariate_boxplot(wine$quality, wine$fixed.acidity,
                                   'fixed acidity'),
             get_bivariate_boxplot(wine$quality, wine$volatile.acidity,
                                   'volatile acidity'),
             get_bivariate_boxplot(wine$quality, wine$citric.acid,
                                   'citric acid'),
             get_bivariate_boxplot(wine$quality, log10(wine$residual.sugar),
                                   'residual sugar'),
             get_bivariate_boxplot(wine$quality, log10(wine$chlorides),
                                   'chlorides'),
             get_bivariate_boxplot(wine$quality, wine$free.sulfur.dioxide,
                                   'free sulf.dioxide'),
             get_bivariate_boxplot(wine$quality, wine$total.sulfur.dioxide,
                                   'total sulf.dioxide'),
             get_bivariate_boxplot(wine$quality, wine$density,
                                   'density'),
             get_bivariate_boxplot(wine$quality, wine$pH,
                                   'pH'),
             get_bivariate_boxplot(wine$quality, log10(wine$sulphates),
                                   'sulphates'),
             get_bivariate_boxplot(wine$quality, wine$alcohol,
                                   'alcohol'),
             ncol = 4)

grid.arrange(get_bivariate_boxplot(wine$rating, wine$fixed.acidity,
                                   'fixed acidity'),
             get_bivariate_boxplot(wine$rating, wine$volatile.acidity,
                                   'volatile acidity'),
             get_bivariate_boxplot(wine$rating, wine$citric.acid,
                                   'citric acid'),
             get_bivariate_boxplot(wine$rating, log10(wine$residual.sugar),
                                   'residual sugar'),
             get_bivariate_boxplot(wine$rating, log10(wine$chlorides),
                                   'chlorides'),
             get_bivariate_boxplot(wine$rating, wine$free.sulfur.dioxide,
                                   'free sulf. dioxide'),
             get_bivariate_boxplot(wine$rating, wine$total.sulfur.dioxide,
                                   'total sulf. dioxide'),
             get_bivariate_boxplot(wine$rating, wine$density,
                                   'density'),
             get_bivariate_boxplot(wine$rating, wine$pH,
                                   'pH'),
             get_bivariate_boxplot(wine$rating, log10(wine$sulphates),
                                   'sulphates'),
             get_bivariate_boxplot(wine$rating, wine$alcohol,
                                   'alcohol'),
             ncol = 4)
			 
```

> The above box plots show that a good wine has the following features:

* higher fixed acidity & citric acid
* lower volatile acidity
* lower pH
* higher sulphates
* higher alcohol

> Let's dig into these features a little more

```{r echo=FALSE, message=FALSE, warning=FALSE, Bivariate_Plots_3}
ggplot(data=wine, aes(x=quality, y=fixed.acidity)) +
  geom_boxplot()
```

> There seems to be very little relation between the quality and \
fixed.acidity.

```{r echo=FALSE, message=FALSE, warning=FALSE, Bivariate_Plots_4}
ggplot(data=wine, aes(x=quality, y=citric.acid)) +
  geom_boxplot()
```

> citric.acid seems to have a strong effect on quality. Quality increases \
with the increase in citric.acid.

```{r echo=FALSE, message=FALSE, warning=FALSE, Bivariate_Plots_5}
ggplot(data=wine, aes(x=quality, y=volatile.acidity)) +
  geom_boxplot()
```

> volatile.acidity also has a strong effect (negative correlation) on quality. \
Quality decreases with the increase in citric.acid.

```{r echo=FALSE, message=FALSE, warning=FALSE, Bivariate_Plots_6}
# Comparison of various acidity variables

ggplot(data = wine, aes(x = fixed.acidity, y = citric.acid)) +
  geom_point()
cor.test(wine$fixed.acidity, wine$citric.acid)

ggplot(data = wine, aes(x = volatile.acidity, y = citric.acid)) +
  geom_point()
cor.test(wine$volatile.acidity, wine$citric.acid)

```

> There is a strong correlation between the acidity variables, as suspected. 

* There is a positive correlation between fixed.acidity and citric.acid \
* There is a negative correlation between volatile.acidity and citric.acid.


```{r echo=FALSE, message=FALSE, warning=FALSE, Bivariate_Plots_7}
# Boxplot for examining chlorides and quality closely
ggplot(data=wine, aes(x=quality, y=chlorides)) +
  geom_boxplot()

```

> There seems to be a weak relation betweeb the two. Lower amount of \
chlorides produces a better wine.

```{r echo=FALSE, message=FALSE, warning=FALSE, Bivariate_Plots_8}
ggplot(data=wine, aes(x=quality, y=pH)) +
  geom_boxplot()
```

> We can see that the better wines have a lower pH value (more acidic) \
though there are a few outliers.


```{r echo=FALSE, message=FALSE, warning=FALSE, Bivariate_Plots_9}
# Examining the sulphate levels
ggplot(data=wine, aes(x=quality, y=sulphates)) +
  geom_boxplot()

```

> There definitely seems to be a trend here even though there are many \
outliers in the medium wines, better wines seem to have a higher \
concentration of sulphates.

```{r echo=FALSE, message=FALSE, warning=FALSE, Bivariate_Plots_10}
#  Alcohol level
ggplot(data=wine, aes(x=quality, y=alcohol)) + 
  geom_boxplot()
```

> We can clearly see the correlation here. As the alcohol level increases \
quality of wine increases with it. But as there are a lot of outliers, let's \
try using a simple linear model to check the connection.

# Bivariate Analysis

### Talk about some of the relationships you observed in this part of the \
investigation. How did the feature(s) of interest vary with other features in \
the dataset?

> Below are some observations:

* Fixed.acidity seems to have little to no effect on quality 
* citric.acid and volatile.acidity have a strong effect on the quality. \
The quality increases with the increase of citric.acid and quality \
decreases with the increase in  volatile.acidity.
* A lower concentration of chlorides seems to produce better wines, even \
though the correlation is weak.
* Better wines are more acidic in nature, but there were many outliers.
* Better wines also seem to have a higher concentration of sulphates.
* Alcohol value has a strong correlation with quality.

### Did you observe any interesting relationships between the other features \
(not the main feature(s) of interest)?

> The various acidity variables are very strongly correlated.

* There is a positive correlation between fixed.acidity and citric.acid
* There is a negative correlation between volatile.acidity and citric.acid.

> The free and total sulphur dioxide variables also seem to have a strong \
positive correlation.

```{r echo=FALSE, message=FALSE, warning=FALSE, Bivariate_Analysis_1}
ggplot(data=wine, aes(x=free.sulfur.dioxide, y=total.sulfur.dioxide)) +
  geom_point()
cor.test(wine$free.sulfur.dioxide, wine$total.sulfur.dioxide)
```

### What was the strongest relationship you found?

> The strongest relationship is between the fixed.acidity, wine$citric.acid \
 with a correlation of 0.6717034. The free and total sulphur dioxide also \
 have a strong relationship with a correlation of 0.6676665.

# Multivariate Plots Section

> Let's create some plots based on the plots of previous sections. The \
features that were highly correlated with quality are plotted below.

```{r echo=FALSE, message=FALSE, warning=FALSE, Multivariate_Plots_1}
ggplot(data = wine, aes(x = alcohol, y = density, color = quality)) +
  geom_point() +
  scale_color_brewer()

ggplot(data = wine,
       aes(x = citric.acid, y = volatile.acidity, color = quality)) +
  geom_point() +
  scale_color_brewer()

ggplot(data = wine,
       aes(x = alcohol, y = sulphates, color = quality)) +
  geom_point() +
  scale_color_brewer()

ggplot(data = wine, aes(x = pH, y = alcohol, color = quality)) +
  geom_point() +
  scale_color_brewer()
```

> The scatter plots look a bit crowded, we can facet these with rating to \
point out the differences between good, average and bad wines.


```{r echo=FALSE, message=FALSE, warning=FALSE, Multivariate_Plots_2}
ggplot(data = wine, aes(x = alcohol, y = density, 
                        color = quality)) +
  geom_point() +
  facet_wrap(~rating) + 
  scale_color_brewer()

ggplot(data = wine, aes(x = citric.acid, y = volatile.acidity, 
                        color = quality)) +
  geom_point() +
  facet_wrap(~rating) + 
  scale_color_brewer()

ggplot(data = wine, aes(x = alcohol, y = sulphates, 
                        color = quality)) +
  geom_point() +
  facet_wrap(~rating) + 
  scale_color_brewer()

ggplot(data = wine, aes(x = pH, y = alcohol, 
                        color = quality)) +
  geom_point() +
  facet_wrap(~rating) + 
  scale_color_brewer()
```

> From the above plots we can tell the following: 

* Density doesn't seem to affect the quality when alcohol is constant.
* Lower volatile.acidity and higher citric acid content make better wines.
* Good wines have a higher sulphate and alcohol content.
* pH doesn't seem to have a huge impact on the quality of wines. Which \
suggests that the quality of wine depends on the type of acids present in it.


# Multivariate Analysis

### Talk about some of the relationships you observed in this part of the \
investigation. Were there features that strengthened each other in terms of \
looking at your feature(s) of interest?
> High alcohol contents and high sulphate concentrations together seem to \
produce better wines.

### Were there any interesting or surprising interactions between features?
> Even though the volatile acidity and citric acid have a considerable \
impact on the quality, pH doesn't play a major role in determining the \
quality of wines.

------

# Final Plots and Summary

### Plot One
```{r echo=FALSE, message=FALSE, warning=FALSE, Final_Plot_1}
ggplot(data = wine, aes(x = quality, y = alcohol, fill = rating)) +
  geom_boxplot() +
  ggtitle('Alcohol Levels for Wines of Different Ratings') +
  xlab('Quality') +
  ylab('Alcohol (% by volume)')
```

### Description One
> These boxplots show how alcohol is a major influence on the quality of wines.

### Plot Two
```{r echo=FALSE, message=FALSE, warning=FALSE, Final_Plot_2}
ggplot(data = wine, aes(x = alcohol, y = sulphates, color = rating)) +
  geom_point() +
  scale_y_continuous(limits=c(0.3, 1.5)) +
  ylab("Potassium Sulphate (g/dm^3)") +
  xlab("Alcohol (% by Volume)") +
  ggtitle("Alcohol and Sulphates over Wine Quality")
```


### Description Two
> Higher alcohol content and higher amount of sulphates combined together \
result in better wines.

### Plot Three

```{r echo=FALSE, message=FALSE, warning=FALSE, Final_Plot_3}
p1 = ggplot(data = wine, aes(x = quality, y = fixed.acidity, 
                             fill = quality)) + 
  xlab('Quality') + 
  ylab('Fixed Acidity (g/dm^3)') +
  geom_boxplot()

p2 = ggplot(data = wine, aes(x = quality, y = volatile.acidity, 
                             fill = quality)) + 
  xlab('Quality') + 
  ylab('Volatile Acidity (g/dm^3)') + 
  geom_boxplot()

p3 = ggplot(data = wine, aes(x = quality, y = citric.acid, 
                             fill = quality)) +
  xlab('Quality') +
  ylab('Citric Acid (g/dm^3)') + 
  geom_boxplot()

p4 = ggplot(data = wine, aes(x = quality, y = pH, 
                             fill = quality)) +
  xlab('Quality') +
  ylab('pH') +
  geom_boxplot()

grid.arrange(p1, p2, p3, p4, ncol = 2)
```

### Description Three
> The above box plots show the effect of various acidity factors on the \
wine quality. Wines of high quality have a low pH value. Citric acid and \
fixed (tartaric) acid have a high positive correlation and low positive \
correlation with the quality of wines respectively. Volatile (acetic) \
acid has a high negative correlation with the quality of wines.

------

# Reflection

> I explored the data by looking at the univariate, bivariate and multivariate \
analysis of the features of the dataset. The features that caught my eye at \
the starting were the alcohol levels and the multiple acidity variables. \
I created an ordered variable from quality to classify the quality into \
ratings. In bivariate analysis, the fixed acidity, volatile acidity, \
alcohol content and sulphate features stood out. To understand the plots \
better, I faceted the plots of important features with rating to get a \
better understanding of the variation in quality. \
One area of improvement that I would suggest for the dataset is to include \
the individual scores of wine critics to better understand their judgement \
as every individual has different taste and rating criteria.