Sumarize and Explore Data by Shashank Verma
========================================================

```{r echo=FALSE, message=FALSE, warning=FALSE, set_directory}
# Set the working directory and list the files
setwd("C:/Users/shank/Desktop/DAND/P4")
list.files()
```

```{r echo=FALSE, message=FALSE, warning=FALSE, packages}
# Load all of the packages that you end up using in your analysis in this code
# chunk.

# Notice that the parameter "echo" was set to FALSE for this code chunk. This
# prevents the code from displaying in the knitted HTML output. You should set
# echo=FALSE for all code chunks in your file, unless it makes sense for your
# report to show the code that generated a particular plot.

# The other parameters for "message" and "warning" should also be set to FALSE
# for other code chunks once you have verified that each plot comes out as you
# want it to. This will clean up the flow of your report.

library(ggplot2)
library(grid)
library(gridExtra)
library(GGally)
library(dplyr)
library(tidyr)
library(reshape)


```

```{r echo=FALSE, Load_the_Data}
# Load the Data
wine <- read.csv("wineQualityReds.csv")
```

> The dataset is related to red variants of the Portuguese "Vinho Verde" wine.
The inputs include objective tests (e.g. PH values) and the output is based on 
sensory data (median of at least 3 evaluations made by wine experts). Each 
expert graded the wine quality between 0 (very bad) and 10 (very excellent).
Only physicochemical (inputs) and sensory (the output) variables are available.

> I will explore the wine quality and physicochemical properties of these wines.
I want to know how the chemical properties influence the quality of red wines.


# Summary Statistics

```{r echo=FALSE, Summary_Statistics}
names(wine)
```

```{r echo=FALSE, Summary_Statistics_2}
str(wine)
```

```{r echo=FALSE, Summary_Statistics_3}
summary(wine)
```

> There are 1599 observations of 13 numeric variables. 
X appears to be the unique identifier. 
quality is an ordered, categorical, discrete variable. From the literature, 
this was on a 0-10 scale, and was rated by at least 3 wine experts. 
The values ranged only from 3 to 8, with a mean of 5.6 and 
median of 6. 
All other variables seem to be continuous quantities (w/ the exception 
of the .sulfur.dioxide suffixes).
From the variable descriptions, it appears that \
fixed.acidity ~ volatile.acidity and \
free.sulfur.dioxide ~ total.sulfur.dioxide may possibly be \
dependent, subsets of each other.


```{r echo=FALSE, Transform_data}

# Transform quality from an integer to an ordered factor
wine$quality <- factor(wine$quality, ordered = T)

# create a rating varible
wine$rating <- ifelse(wine$quality < 5, 'bad', 
                      ifelse(wine$quality < 7, 'average', 'good'))
wine$rating <- ordered(wine$rating, levels = c('bad', 'average', 'good'))
```

# Univariate Plots Section

> Now, I'll draw quick histograms of all the features for showing the \
distributions of the variables.

```{r echo=FALSE, Univariate_Plots_1}
fix_ac_plot = ggplot(aes(x = fixed.acidity), data = wine) +
  geom_histogram(stat = "count")

vol_ac_plot = ggplot(aes(x = volatile.acidity), data = wine) +
  geom_histogram(stat = "count")

cit_ac_plot = ggplot(aes(x = citric.acid), data = wine) +
  geom_histogram(stat = "count")

res_sug_plot = ggplot(aes(x = residual.sugar), data = wine) +
  geom_histogram(stat = "count")

chlor_plot = ggplot(aes(x = chlorides), data = wine) +
  geom_histogram(stat = "count")

free_sul_plot = ggplot(aes(x = free.sulfur.dioxide), data = wine) +
  geom_histogram(stat = "count")

tot_sul_plot = ggplot(aes(x = total.sulfur.dioxide), data = wine) +
  geom_histogram(stat = "count")

den_plot = ggplot(aes(x = density), data = wine) +
  geom_histogram(stat = "count")

ph_plot = ggplot(aes(x = pH), data = wine) +
  geom_histogram(stat = "count")

sul_plot = ggplot(aes(x = sulphates), data = wine) +
  geom_histogram(stat = "count")

al_plot = ggplot(aes(x = alcohol), data = wine) +
  geom_histogram(stat = "count")

qual_plot = ggplot(aes(x = quality), data = wine) +
  geom_histogram(stat = "count")

grid.arrange(fix_ac_plot, vol_ac_plot, cit_ac_plot, res_sug_plot, 
chlor_plot, free_sul_plot, tot_sul_plot, den_plot, ph_plot, 
sul_plot, al_plot, qual_plot, ncol = 3)

```

> pH is normally distributed, but fixed.acidity and citric.acid appear to \
be positively skewed. We will draw log 10 plots of these variables to \
analyse further.

> Here are some quick box plots for all the numerical variables. We can \
visualize the outliers here.
 
```{r echo=FALSE, Univariate_Plots_2}
meltData <- melt(wine[2:13])
ggplot(meltData, aes(factor(variable), value))+ 
  geom_boxplot() + facet_wrap(~variable, scale="free")
```

```{r echo=FALSE, Univariate_Plots_3}
ggplot(wine, aes(x = fixed.acidity)) +
  geom_histogram() +
  scale_x_log10()

ggplot(wine, aes(x = volatile.acidity)) +
  geom_histogram() +
  scale_x_log10()

ggplot(wine, aes(x = citric.acid)) +
  geom_histogram() +
  scale_x_log10()
```
> On a log 10 scale, fixed.acidity and volatile.acidity appear to have a \
normal distribution, which makes sense since pH has a normal distribution. \
However, citric.acid does not appear to have a normal distribution.

```{r}
length(subset(wine, citric.acid == 0)$citric.acid)
```

> After checking further, we find that the citric.acid contains 132 zero \
values. This can arise if the values were not entered.

```{r}
p1 <- ggplot(aes(x = residual.sugar), data = wine) +
  geom_histogram()

p2 <- p1 + 
  scale_x_continuous(lim = c(0, quantile(wine$residual.sugar, 0.95))) +
  xlab('residual.sugar, 95th percentile truncated')

p3 <- p2 + scale_x_log10() + xlab('residual.sugar, log10')
grid.arrange(p1, p2, p3, ncol=1)

p1 <- ggplot(aes(x = chlorides), data = wine) +
  geom_histogram()

p2 <- p1 + 
  scale_x_continuous(lim = c(0, quantile(wine$chlorides, 0.95))) +
  xlab('chlorides, 95th percentile truncated')

p3 <- p2 + scale_x_log10() + xlab('chlorides, log10')
grid.arrange(p1, p2, p3, ncol=1)

p1 <- ggplot(aes(x = sulphates), data = wine) +
  geom_histogram()

p2 <- p1 + 
  scale_x_continuous(lim = c(0, quantile(wine$sulphates, 0.95))) +
  xlab('sulphates, 95th percentile truncated')

p3 <- p2 + scale_x_log10() + xlab('sulphates, log10')
grid.arrange(p1, p2, p3, ncol=1)

```

> Above are the plots for residual.sugar, chlorides and sulphates with \
the original plot, 95th percentile truncated and then the log 10 plots.
We can get a better understanding of the distribution here.

# Univariate Analysis

> Below are the distributions of the features found in our dataset:
* Positively Skewed: fixed.acidity, citric.acid, free.sulphur.dioxide, \
total.sulphur.dioxide, sulphates, alcohol
* Normal: volatile.acidity, density, pH
* Long Tailed: residual.sugar, chlorides


### What is the structure of your dataset?
> The dataset consists of 1599 observations of 13 variables. quality is an 
ordered categorical variable and all the other variables are numerical.

### What is/are the main feature(s) of interest in your dataset?
> The main feature of interest here is the quality of wine. It ranges from 
3 to 8 in the dataset.

### What other features in the dataset do you think will help support your \
investigation into your feature(s) of interest?
> The multiple acidity variables can be related to each other and the quality \
of the wines. The alcohol and residual.sugar variables also look interesting.


### Did you create any new variables from existing variables in the dataset?
> I transformed the quality from an integer to an ordered factor and created \
a variable called rating to classify the wines as good, average or bad.


### Of the features you investigated, were there any unusual distributions? \
Did you perform any operations on the data to tidy, adjust, or change the form \
of the data? If so, why did you do this?
> I did not perform any operations to tidy the data, only changed the \
quality from an integer to an ordered factor and created a variable \
called rating to classify the wines as good, average or bad as this will help \
in better visualising the quality of wines.

# Bivariate Plots Section

> **Tip**: Based on what you saw in the univariate plots, what relationships
between variables might be interesting to look at in this section? Don't limit
yourself to relationships between a main output feature and one of the
supporting variables. Try to look at relationships between supporting variables
as well.

```{r echo=FALSE, Bivariate_Cors}
cor(
  wine %>%
    # first we remove unwanted columns
    dplyr::select(-X) %>%
    dplyr::select(-rating) %>%
    mutate(
      # now we translate quality to a number
      quality = as.numeric(quality)
    )
  )

```

> From the above correlations we can see that the quality is correlated with \
* alcohol (positive correlation)
* volatile.acidity (negative correlation)
* sulphates (positive correlation)
* citric.acid (positive correlation)

> Now we'll create scatterplots of these variables with each other with \
rating as facet.

```{r}
ggplot(data = wine, aes(x = log10(sulphates), y = alcohol)) +
  facet_wrap(~rating) +
  geom_point()

ggplot(data = wine, aes(x = volatile.acidity, y = alcohol)) +
  facet_wrap(~rating) +
  geom_point()

ggplot(data = wine, aes(x = citric.acid, y = alcohol)) +
  facet_wrap(~rating) +
  geom_point()

ggplot(data = wine, aes(x = volatile.acidity, y = log10(sulphates))) +
  facet_wrap(~rating) +
  geom_point()

ggplot(data = wine, aes(x = citric.acid, y = log10(sulphates))) +
  facet_wrap(~rating) +
  geom_point()

ggplot(data = wine, aes(x = citric.acid, y = volatile.acidity)) +
  facet_wrap(~rating) +
  geom_point()
```


```{r}
get_bivariate_boxplot <- function(x, y, ylab) {
  return(ggplot(data = wine, aes(x = x, y = y)) + geom_boxplot() + ylab(ylab))
}

grid.arrange(get_bivariate_boxplot(wine$quality, wine$fixed.acidity,
                                   'fixed acidity'),
             get_bivariate_boxplot(wine$quality, wine$volatile.acidity,
                                   'volatile acidity'),
             get_bivariate_boxplot(wine$quality, wine$citric.acid,
                                   'citric acid'),
             get_bivariate_boxplot(wine$quality, log10(wine$residual.sugar),
                                   'residual sugar'),
             get_bivariate_boxplot(wine$quality, log10(wine$chlorides),
                                   'chlorides'),
             get_bivariate_boxplot(wine$quality, wine$free.sulfur.dioxide,
                                   'free sulf.dioxide'),
             get_bivariate_boxplot(wine$quality, wine$total.sulfur.dioxide,
                                   'total sulf.dioxide'),
             get_bivariate_boxplot(wine$quality, wine$density,
                                   'density'),
             get_bivariate_boxplot(wine$quality, wine$pH,
                                   'pH'),
             get_bivariate_boxplot(wine$quality, log10(wine$sulphates),
                                   'sulphates'),
             get_bivariate_boxplot(wine$quality, wine$alcohol,
                                   'alcohol'),
             ncol = 4)

grid.arrange(get_bivariate_boxplot(wine$rating, wine$fixed.acidity,
                                   'fixed acidity'),
             get_bivariate_boxplot(wine$rating, wine$volatile.acidity,
                                   'volatile acidity'),
             get_bivariate_boxplot(wine$rating, wine$citric.acid,
                                   'citric acid'),
             get_bivariate_boxplot(wine$rating, log10(wine$residual.sugar),
                                   'residual sugar'),
             get_bivariate_boxplot(wine$rating, log10(wine$chlorides),
                                   'chlorides'),
             get_bivariate_boxplot(wine$rating, wine$free.sulfur.dioxide,
                                   'free sulf. dioxide'),
             get_bivariate_boxplot(wine$rating, wine$total.sulfur.dioxide,
                                   'total sulf. dioxide'),
             get_bivariate_boxplot(wine$rating, wine$density,
                                   'density'),
             get_bivariate_boxplot(wine$rating, wine$pH,
                                   'pH'),
             get_bivariate_boxplot(wine$rating, log10(wine$sulphates),
                                   'sulphates'),
             get_bivariate_boxplot(wine$rating, wine$alcohol,
                                   'alcohol'),
             ncol = 4)
			 
```

> The above box plots show that a good wine has the following features:\
* higher fixed acidity & citric acid
* lower volatile acidity
* lower pH
* higher sulphates
* higher alcohol

> Let's dig into these features a little more

```{r}
ggplot(data=wine, aes(x=quality, y=fixed.acidity)) +
  geom_boxplot()
```

> There seems to be very little relation between the quality and \
fixed.acidity.

```{r}
ggplot(data=wine, aes(x=quality, y=citric.acid)) +
  geom_boxplot()
```

> citric.acid seems to have a strong effect on quality. Quality increases \
with the increase in citric.acid.

```{r}
ggplot(data=wine, aes(x=quality, y=volatile.acidity)) +
  geom_boxplot()
```

> volatile.acidity also has a strong effect (negative correlation) on quality. \
Quality decreases with the increase in citric.acid.

```{r}
# Comparison of various acidity variables

ggplot(data = wine, aes(x = fixed.acidity, y = citric.acid)) +
  geom_point()
cor.test(wine$fixed.acidity, wine$citric.acid)

ggplot(data = wine, aes(x = volatile.acidity, y = citric.acid)) +
  geom_point()
cor.test(wine$volatile.acidity, wine$citric.acid)

```

> There is a strong correlation between the acidity variables, as suspected. \
* There is a positive correlation between fixed.acidity and citric.acid \
* There is a negative correlation between volatile.acidity and citric.acid.


```{r}
# Boxplot for examining chlorides and quality closely
ggplot(data=wine, aes(y=chlorides, x=quality)) +
  geom_boxplot()

```

> There seems to be a weak relation betweeb the two. Lower amount of \
chlorides produces a better wine.

```{r}
ggplot(data=wine, aes(x=quality, y=pH)) +
  geom_boxplot()
```

> We can see that the better wines have a lower pH value (more acidic) \
though there are a few outliers.


```{r}
# Examining the sulphate levels
ggplot(data=wine, aes(y=sulphates, x=quality)) +
  geom_boxplot()

```

> There definitely seems to be a trend here even though there are many \
outliers in the medium wines, better wines seem to have a higher \
concentration of sulphates.

```{r}
#  Alcohol level
ggplot(data=wine, aes(y=alcohol, x=quality)) + 
  geom_boxplot()
```

> We can clearly see the correlation here. As the alcohol level increases \
quality of wine increases with it. But as there are a lot of outliers, let's \
try using a simple linear model to check the connection.

```{r}
alcoholQual <- lm(as.numeric(quality) ~ alcohol, data = wine)
summary(alcoholQual)
alcoholDF = data.frame(wine$quality )
alcoholDF$predictions <- predict(alcoholQual, wine)
alcoholDF$error <- (alcoholDF$predictions - as.numeric(wine$quality))/as.numeric(wine$quality)

ggplot(data=alcoholDF, aes(x=wine.quality, y=error)) +
  geom_boxplot()
```

> We get an R-squared value of 0.2263. Which is not enough by itself. Let's \
look at other variables to generate a better model.


# Bivariate Analysis

### Talk about some of the relationships you observed in this part of the \ 
investigation. How did the feature(s) of interest vary with other features in \
the dataset?

> *Fixed.acidity seems to have little to no effect on quality \
* citric.acid and volatile.acidity have a strong effect on the quality. \
The quality increases with the increase of citric.acid and quality \
decreases with the increase in  volatile.acidity.
* A lower concentration of chlorides seems to produce better wines, even \
though the correlation is weak.
* Better wines are more acidic in nature, but there were many outliers.
* Better wines also seem to have a higher concentration of sulphates.
* Alcohol value has a strong correlation with quality, but we can't explain \
this with variance alone. We need to look at more variables to get a \
better model.

### Did you observe any interesting relationships between the other features \
(not the main feature(s) of interest)?

> The various acidity variables are very strongly correlated.
* There is a positive correlation between fixed.acidity and citric.acid
* There is a negative correlation between volatile.acidity and citric.acid.

> The free and total sulphur dioxide variables also seem to have a strong \
positive correlation.

```{r}
ggplot(data=wine, aes(x=free.sulfur.dioxide, y=total.sulfur.dioxide)) +
  geom_point()
cor.test(wine$free.sulfur.dioxide, wine$total.sulfur.dioxide)
```

### What was the strongest relationship you found?

> The strongest relationship is between the fixed.acidity, wine$citric.acid \
 with a correlation of 0.6717034. The free and total sulphur dioxide also \
 have a strong relationship with a correlation of 0.6676665.

# Multivariate Plots Section

> **Tip**: Now it's time to put everything together. Based on what you found in
the bivariate plots section, create a few multivariate plots to investigate
more complex interactions between variables. Make sure that the plots that you
create here are justified by the plots you explored in the previous section. If
you plan on creating any mathematical models, this is the section where you
will do that.

```{r echo=FALSE, Multivariate_Plots_1}
# Density of Quality by color

```


# Multivariate Analysis

### Talk about some of the relationships you observed in this part of the \
investigation. Were there features that strengthened each other in terms of \
looking at your feature(s) of interest?

### Were there any interesting or surprising interactions between features?

### OPTIONAL: Did you create any models with your dataset? Discuss the \
strengths and limitations of your model.

------

# Final Plots and Summary

> **Tip**: You've done a lot of exploration and have built up an understanding
of the structure of and relationships between the variables in your dataset.
Here, you will select three plots from all of your previous exploration to
present here as a summary of some of your most interesting findings. Make sure
that you have refined your selected plots for good titling, axis labels (with
units), and good aesthetic choices (e.g. color, transparency). After each plot,
make sure you justify why you chose each plot by describing what it shows.

### Plot One
```{r echo=FALSE, Plot_One}

```

### Description One


### Plot Two
```{r echo=FALSE, Plot_Two}

```

### Description Two


### Plot Three
```{r echo=FALSE, Plot_Three}

```

### Description Three

------

# Reflection

> **Tip**: Here's the final step! Reflect on the exploration you performed and
the insights you found. What were some of the struggles that you went through?
What went well? What was surprising? Make sure you include an insight into
future work that could be done with the dataset.

> **Tip**: Don't forget to remove this, and the other **Tip** sections before
saving your final work and knitting the final report!