Sumarize and Explore Data by Shashank Verma
========================================================

```{r echo=FALSE, message=FALSE, warning=FALSE, set_directory}
# Set the working directory and list the files
setwd("C:/Users/shank/Desktop/DAND/P4")
list.files()
```

```{r echo=FALSE, message=FALSE, warning=FALSE, packages}
# Load all of the packages that you end up using in your analysis in this code
# chunk.

# Notice that the parameter "echo" was set to FALSE for this code chunk. This
# prevents the code from displaying in the knitted HTML output. You should set
# echo=FALSE for all code chunks in your file, unless it makes sense for your
# report to show the code that generated a particular plot.

# The other parameters for "message" and "warning" should also be set to FALSE
# for other code chunks once you have verified that each plot comes out as you
# want it to. This will clean up the flow of your report.

library(ggplot2)
library(grid)
library(gridExtra)
library(GGally)
library(dplyr)
library(tidyr)
library(reshape)

```

```{r echo=FALSE, Load_the_Data}
# Load the Data
wine <- read.csv("wineQualityReds.csv")
```

> The dataset is related to red variants of the Portuguese "Vinho Verde" wine.
The inputs include objective tests (e.g. PH values) and the output is based on 
sensory data (median of at least 3 evaluations made by wine experts). Each 
expert graded the wine quality between 0 (very bad) and 10 (very excellent).
Only physicochemical (inputs) and sensory (the output) variables are available.

> I will explore the wine quality and physicochemical properties of these wines.
I want to know how the chemical properties influence the quality of red wines.


# Summary Statistics

```{r echo=FALSE, Summary_Statistics}
names(wine)
```

```{r echo=FALSE, Summary_Statistics_2}
str(wine)
```

```{r echo=FALSE, Summary_Statistics_3}
summary(wine)
```

> There are 1599 observations of 13 numeric variables. 
X appears to be the unique identifier. 
quality is an ordered, categorical, discrete variable. From the literature, 
this was on a 0-10 scale, and was rated by at least 3 wine experts. 
The values ranged only from 3 to 8, with a mean of 5.6 and 
median of 6. 
All other variables seem to be continuous quantities (w/ the exception 
of the .sulfur.dioxide suffixes).
From the variable descriptions, it appears that \
fixed.acidity ~ volatile.acidity and \
free.sulfur.dioxide ~ total.sulfur.dioxide may possibly be \
dependent, subsets of each other.


```{r echo=FALSE, Transform_data}

# Transform quality from an integer to an ordered factor
wine$quality <- factor(wine$quality, ordered = T)

# create a rating varible
wine$rating <- ifelse(wine$quality < 5, 'bad', 
                      ifelse(wine$quality < 7, 'average', 'good'))
wine$rating <- ordered(wine$rating, levels = c('bad', 'average', 'good'))
```

# Univariate Plots Section

> Now, I'll draw quick histograms of all the features for showing the \
distributions of the variables.

```{r echo=FALSE, Univariate_Plots_2}
fix_ac_plot = ggplot(aes(x = fixed.acidity), data = wine) +
  geom_histogram(stat = "count")

vol_ac_plot = ggplot(aes(x = volatile.acidity), data = wine) +
  geom_histogram(stat = "count")

cit_ac_plot = ggplot(aes(x = citric.acid), data = wine) +
  geom_histogram(stat = "count")

res_sug_plot = ggplot(aes(x = residual.sugar), data = wine) +
  geom_histogram(stat = "count")

chlor_plot = ggplot(aes(x = chlorides), data = wine) +
  geom_histogram(stat = "count")

free_sul_plot = ggplot(aes(x = free.sulfur.dioxide), data = wine) +
  geom_histogram(stat = "count")

tot_sul_plot = ggplot(aes(x = total.sulfur.dioxide), data = wine) +
  geom_histogram(stat = "count")

den_plot = ggplot(aes(x = density), data = wine) +
  geom_histogram(stat = "count")

ph_plot = ggplot(aes(x = pH), data = wine) +
  geom_histogram(stat = "count")

sul_plot = ggplot(aes(x = sulphates), data = wine) +
  geom_histogram(stat = "count")

al_plot = ggplot(aes(x = alcohol), data = wine) +
  geom_histogram(stat = "count")

qual_plot = ggplot(aes(x = quality), data = wine) +
  geom_histogram(stat = "count")

grid.arrange(fix_ac_plot, vol_ac_plot, cit_ac_plot, res_sug_plot, chlor_plot, free_sul_plot, tot_sul_plot, den_plot, ph_plot, sul_plot, al_plot, qual_plot, ncol = 3)

```

> Here are some quick box plots for all the numerical variables. We can see the
 number of outliers.
 
```{r echo=FALSE, Univariate_Plots_1}
meltData <- melt(wine[2:13])
ggplot(meltData, aes(factor(variable), value))+ 
  geom_boxplot() + facet_wrap(~variable, scale="free")
```

# Univariate Analysis

> Below are the distributions of the features found in our dataset:
* Positively Skewed: fixed.acidity, citric.acid, free.sulphur.dioxide, \
total.sulphur.dioxide, sulphates, alcohol
* Normal: volatile.acidity, density, pH
* Long Tailed: residual.sugar, chlorides


### What is the structure of your dataset?
> The dataset consists of 1599 observations of 13 variables. quality is an 
ordered categorical variable and all the other variables are numerical.

### What is/are the main feature(s) of interest in your dataset?
> The main feature of interest here is the quality of wine. It ranges from 
3 to 8 in the dataset.

### What other features in the dataset do you think will help support your \
investigation into your feature(s) of interest?
> The multiple acidity variables can be related to each other and the quality \
of the wines. The alcohol and residual.sugar variables also look interesting.


### Did you create any new variables from existing variables in the dataset?
> I transformed the quality from an integer to an ordered factor and created \
a variable called rating to classify the wines as good, average or bad.


### Of the features you investigated, were there any unusual distributions? \
Did you perform any operations on the data to tidy, adjust, or change the form \
of the data? If so, why did you do this?
> I did not perform any operations to tidy the data, only changed the \
quality from an integer to an ordered factor and created a variable \
called rating to classify the wines as good, average or bad as this will help \
in better visualising the quality of wines.

# Bivariate Plots Section

> **Tip**: Based on what you saw in the univariate plots, what relationships
between variables might be interesting to look at in this section? Don't limit
yourself to relationships between a main output feature and one of the
supporting variables. Try to look at relationships between supporting variables
as well.

```{r echo=FALSE, Bivariate_Plots}
#wine2 <- wine
#wine2$quality <- as.numeric(as.character(wine$quality))
#cor(wine2)

cor(
  wine %>%
    # first we remove unwanted columns
    dplyr::select(-X) %>%
    dplyr::select(-rating) %>%
    mutate(
      # now we translate quality to a number
      quality = as.numeric(quality)
    )
  )

#emphasize.strong.cells(which(abs(c) > .3 & c != 1, arr.ind = TRUE))
#pandoc.table(c)
```

# Bivariate Analysis

> **Tip**: As before, summarize what you found in your bivariate explorations
here. Use the questions below to guide your discussion.

### Talk about some of the relationships you observed in this part of the \
investigation. How did the feature(s) of interest vary with other features in \
the dataset?

### Did you observe any interesting relationships between the other features \
(not the main feature(s) of interest)?

### What was the strongest relationship you found?


# Multivariate Plots Section

> **Tip**: Now it's time to put everything together. Based on what you found in
the bivariate plots section, create a few multivariate plots to investigate
more complex interactions between variables. Make sure that the plots that you
create here are justified by the plots you explored in the previous section. If
you plan on creating any mathematical models, this is the section where you
will do that.

```{r echo=FALSE, Multivariate_Plots_1}
# Density of Quality by color
qplot(quality, data = data, binwidth = 1, color = color, geom = "density") + 
    scale_x_continuous(breaks = seq(3, 9, 1))

ggplot()
```




# Multivariate Analysis




### Talk about some of the relationships you observed in this part of the \
investigation. Were there features that strengthened each other in terms of \
looking at your feature(s) of interest?

### Were there any interesting or surprising interactions between features?

### OPTIONAL: Did you create any models with your dataset? Discuss the \
strengths and limitations of your model.

------

# Final Plots and Summary

> **Tip**: You've done a lot of exploration and have built up an understanding
of the structure of and relationships between the variables in your dataset.
Here, you will select three plots from all of your previous exploration to
present here as a summary of some of your most interesting findings. Make sure
that you have refined your selected plots for good titling, axis labels (with
units), and good aesthetic choices (e.g. color, transparency). After each plot,
make sure you justify why you chose each plot by describing what it shows.

### Plot One
```{r echo=FALSE, Plot_One}

```

### Description One


### Plot Two
```{r echo=FALSE, Plot_Two}

```

### Description Two


### Plot Three
```{r echo=FALSE, Plot_Three}

```

### Description Three

------

# Reflection

> **Tip**: Here's the final step! Reflect on the exploration you performed and
the insights you found. What were some of the struggles that you went through?
What went well? What was surprising? Make sure you include an insight into
future work that could be done with the dataset.

> **Tip**: Don't forget to remove this, and the other **Tip** sections before
saving your final work and knitting the final report!